---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
  
  
**Code to read in and process the data:**
```{r processData}
options(scipen=999)
library(lubridate)
# setwd("RepData_PeerAssessment1")
unzip("activity.zip","activity.csv")
activity <- read.csv(file = "activity.csv")
activity$date <- ymd(activity$date)
```


## What is mean total number of steps taken per day?
```{r plotTotalSteps}
library(ggplot2)
suppressPackageStartupMessages(library(dplyr,quietly = TRUE))
stepsPerDay <- activity %>% 
  group_by(date) %>% 
  summarize("total" = sum(steps))
# stepsPerDay <- activity %>% group_by(date) %>% summarize("total" = sum(steps,na.rm=TRUE))
meanStepsPerDay   <- mean(stepsPerDay$total,na.rm=TRUE)
medianStepsPerDay <- median(stepsPerDay$total,na.rm=TRUE)
ggplot(stepsPerDay,aes(x=total))+
  geom_histogram()+
  geom_vline(aes(xintercept=meanStepsPerDay,linetype="Mean",color="Mean"),show_guide=TRUE)+
  geom_vline(aes(xintercept=medianStepsPerDay,linetype="Median",color="Median"),show_guide=TRUE)+
  scale_linetype_manual(name="Central Values", values=c(Mean="solid",Median="dashed"), guide=FALSE)+
  scale_color_manual(name="Central Values",values=c(Mean="red",Median="cyan"))+
  xlab("Steps Per Day")+
  ggtitle("Histogram of Total Steps Taken Each Day")
c(meanStepsPerDay,medianStepsPerDay,sum(stepsPerDay$total,na.rm=TRUE))
```

**Mean** steps per day: `r meanStepsPerDay`  
**Median** steps per day: `r medianStepsPerDay`  
**Total** steps: `r sum(stepsPerDay$total,na.rm=TRUE)`


## What is the average daily activity pattern?

```{r plotStepsPerInterval}
stepsPerInterval <- activity %>% 
  group_by(interval) %>% 
  summarize("average" = mean(steps,na.rm=TRUE))
ggplot(stepsPerInterval,aes(x=interval,y=average))+
  geom_line()+
  xlab("5 Minute Interval")+
  ylab("Mean Steps per 5 Minute Interval")+
  ggtitle("Average Number of Steps Taken (Averaged Across All Days)")
maxAverageInterval <- arrange(stepsPerInterval,desc(average))
maxAverageInterval[1,]
```

The **five-minute interval with largest number of average steps** is number: `r maxAverageInterval[1,1]`.  
This interval had an average of `r maxAverageInterval[1,2]` steps.  

## Imputing missing values

```{r computeNAs}
numRowsWithNA <- sum(!complete.cases(activity))
numRowsWithNA
```

There are `r numRowsWithNA` rows with NAs in the original data.  

**Strategy for imputing missing data:**  
Each row with missing data will have data imputed. The imputed value will equal the global average for that 5-minute interval on that day of the week.

**Code for imputing missing data:**
```{r imputeMissingData}
imputedActivity <- activity %>% 
  group_by(wday(date)) %>% 
  mutate(dailyAverage = mean(steps,na.rm = TRUE))
imputedActivity[!complete.cases(imputedActivity),"steps"] <- 
  imputedActivity[!complete.cases(imputedActivity),"dailyAverage"]
imputedActivity <- imputedActivity[,c("steps","date","interval")]
imputedStepsPerDay <- imputedActivity %>% 
  group_by(date) %>% 
  summarize("total" = sum(steps))
imputedMeanStepsPerDay   <- mean(imputedStepsPerDay$total)
imputedMedianStepsPerDay <- median(imputedStepsPerDay$total)
```

### The Impact of Imputing Missing Values

```{r plotTotalStepsImputed}
ggplot(imputedStepsPerDay,aes(x=total))+
  geom_histogram()+
  geom_vline(aes(xintercept=imputedMeanStepsPerDay,linetype="Mean",color="Mean"),show_guide=TRUE)+
  geom_vline(aes(xintercept=imputedMedianStepsPerDay,linetype="Median",color="Median"),show_guide=TRUE)+
  scale_linetype_manual(name="Central Values", values=c(Mean="solid",Median="dashed"), guide=FALSE)+
  scale_color_manual(name="Central Values",values=c(Mean="red",Median="cyan"))+
  xlab("Steps Per Day")

c(imputedMeanStepsPerDay,imputedMedianStepsPerDay,sum(imputedStepsPerDay$total))
```

Imputed **Mean** steps per day: `r imputedMeanStepsPerDay`  
Imputed **Median** steps per day: `r imputedMedianStepsPerDay`  
Imputed **Total** steps: `r sum(imputedStepsPerDay$total)`

```{r impactOfImputation}
fracOriginallyMissing <- sum(is.na(activity$steps))/nrow(activity)
fracImputedIncrease   <- (sum(imputedStepsPerDay$total)-sum(stepsPerDay$total,na.rm=TRUE))/sum(stepsPerDay$total,na.rm=TRUE)
c(paste(round(100*fracImputedIncrease, 2),"%", sep=""),paste(round(100*fracOriginallyMissing, 2), "%", sep=""))
```

Total step count has increased `r paste(round(100*fracImputedIncrease, 2), "%", sep="")`, which is in-line with expectation, as `r paste(round(100*fracOriginallyMissing, 2), "%", sep="")` rows of rows were imputed.  

The slight increase in mean and median step counts per day is because missing values were disproportionately imputed more from Mondays and Fridays, and less from Tuesdays.

```{r computeDailyImputationProportion}
activity %>% 
  group_by(wday(date,label = TRUE)) %>% 
  summarise(fracNAs = sum(is.na(steps))) %>% 
  mutate(fracNAs = paste(round(100*fracNAs/sum(fracNAs), 2),"%",sep=""))
```

## Are there differences in activity patterns between weekdays and weekends?

```{r plotStepsPerIntervalPanels}
factoredImputedActivity <- 
  imputedActivity %>% 
  mutate(wday = (wday(date))) %>% 
  mutate(wday = (wday>1 & wday<7)) %>% 
  mutate(wday = factor(wday,labels=c("weekend","weekday")))

averagedByWeekdayImputedActivity <- 
  factoredImputedActivity %>% 
  group_by(wday,interval) %>% 
  summarize(meanStepsPerInterval=mean(steps))

ggplot(averagedByWeekdayImputedActivity, 
       aes(x=interval,y=meanStepsPerInterval))+
  geom_line()+
  facet_wrap(~wday,nrow=2)+
  xlab("5 Minute Interval")+
  ylab("Mean Steps Per 5 Minute Interval")
```
